%% ============================================================================
%% FULL PAPER — ASSEMBLED DRAFT (LaTeX, 2026-06-08) — Overleaf-ready.
%% Generated from paper_workspace/drafts/*.md. Compiles with pdfLaTeX (elsarticle).
%% Open items are rendered in red via \todo{...} so they are visible and the file
%% still compiles. Source of truth for prose: drafts/ ; for numbers: 00_design/G1.
%% Section cross-references use \ref{} (auto-numbered), so LaTeX numbering is correct
%% even though Data/Results/Discussion shift vs the markdown's provisional "5b" scheme.
%% ============================================================================
\documentclass[review]{elsarticle}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{siunitx}
\usepackage{xcolor}

% Visible placeholder for anything still pending (G5 lit, tables, figures, abstract).
\newcommand{\todo}[1]{{\color{red}\textbf{[#1]}}}
\newcommand{\fcr}{\mathrm{FCR}}

\journal{Transportation Research Part C: Emerging Technologies}

% Suppress the elsarticle "Preprint submitted to ... <date>" footer.
\makeatletter
\def\ps@pprintTitle{%
  \let\@oddhead\@empty
  \let\@evenhead\@empty
  \def\@oddfoot{}%
  \let\@evenfoot\@oddfoot}
\makeatother

\begin{document}

\begin{frontmatter}

\title{Maritime Speed Optimization Under Weather Uncertainty}

\author[tau]{Ami Shafir}
\author[tau]{Tal Raviv}
\address[tau]{Faculty of Engineering, Tel Aviv University, Tel Aviv, Israel}

\begin{abstract}
\todo{To be written.}
\end{abstract}

\begin{keyword}
ship speed optimization \sep dynamic programming \sep rolling horizon \sep fuel consumption \sep
weather routing \sep Jensen's inequality
\end{keyword}

\end{frontmatter}

%% ============================================================ 1. INTRODUCTION
\section{Introduction}
\label{sec:intro}
\todo{To be written.}

%% ============================================================ 2. RELATED WORK
\section{Related work}
\label{sec:related}
\todo{To be written.}

%% ============================================================ 3. PROBLEM FORMULATION
\section{Problem formulation}
\label{sec:problem}

In this paper we study the problem of controling the speed of sea vessels with the goal of minimizing it total voyage fule consumption considering dynamic and stochastic sea conditions and utlizing noisy weather forecasts.  As a first step, we define a dynamic determinstic version of the problem were the sea conditions at each location along the plan route at each time during the planning horizon  is known and we then extened to a stocastic dynamic version where a noisy forecast is given and dynamically updated throughout the voyage.  In this section we describe the input of the problem and the method by which the fuel consuption is calculated. This description requires some understanding of the physical asspect of sea voyage which we berifely provide here. 

\begin{description}
	\item[Route] A route is defined by a set by sequence of geographical waypoints assuming the vessel starts its route at the first waypoint, travels on a great circle between each pair of consecutive way points in the sequence.   The waypoints are  indexed $0,\ldots, N$ and each pair of waypoints define a \emph{segment} numbered $1,\ldots, N$.
	\item[T]  Total required voyage time (in hours),  commonly refred to in the literature as expected time of arrival (ETA).
	\item [Weather]    The weather condition are given by the wind speed and direciton, sea current speed and direction, and wave condition measured in Beufor units. In reality these conditions changes continuesly in time and space but our model they are assume fixed for ea....
	\item[FCR]   Fuel consumption rate (in metric ton of bunker fuel per hour). The FCR is calculated as a function of the weather conditions, the physical caracteristics of the vessle, the course (sailing direction)  and the speed over ground (SOG)  of the vessel at each particular time. The calculation or estimation of this function is an independent challange which is not part of the contribution of this paper. We use a method propsoed by (CITE Sustenability) which is described in Appnedix X. We assume that the vessel charactersitcs are fixed, .  The vessle course is fixed at of the route segments and the SOG is a decision variable of the speed optinization problem.
	
\end{description}




\subsection{Vessel and voyage}
The vessel modelled in this study is a representative oil products tanker with the principal
particulars in Table~\ref{tab:ship}. A voyage is
a fixed route of total length $L$ to be completed within a required arrival time $T$ (the ETA). The
route is partitioned into legs $i=1,\dots,N$ of length $d_i$, with weather specified per leg and
time. The control problem is to choose a speed profile that minimises total fuel while arriving
within $T$.

\begin{table}[ht]
\centering
\caption{Principal particulars of the representative oil products tanker; resistance model and
coefficient tables from \citet{yang2020}.}
\label{tab:ship}
\begin{tabular}{lll}
\toprule
Parameter & Symbol & Value \\
\midrule
Length between perpendiculars & $L_{pp}$ & \SI{200}{\meter} \\
Beam & $B$ & \SI{32}{\meter} \\
Draft & --- & \SI{12}{\meter} \\
Displacement & $\Delta$ & \SI{50000}{\tonne} \\
Block coefficient & $C_b$ & 0.75 \\
Installed power & --- & \SI{10000}{\kilo\watt} \\
Still-water speed range & $V_s$ & 11--13 kn \\
\bottomrule
\end{tabular}
\end{table}

Two routes are studied, chosen to contrast weather regimes: Route~1 (Persian Gulf $\rightarrow$
Strait of Malacca, $L=3{,}393$ nm, $T=\SI{280}{\hour}$), in mild and relatively uniform conditions,
and Route~2 (St.\ John's $\rightarrow$ Liverpool, North Atlantic, $L=1{,}955$ nm,
$T=\SI{168}{\hour}$), in harsher and more variable conditions. Full route and weather statistics are
given in Section~\ref{sec:data}.

\subsection{Speed over ground}
\label{sec:sog}
The vessel's still-water speed (SWS), $V_s$, is the speed it would make in calm water at a given
power setting. The speed actually achieved over the ground (SOG), $V_g$, is $V_s$ modified by the
environment: wind and waves reduce it, and the along-track component of the current adds to or
subtracts from it. This study uses the resistance-decomposition model of \citet{yang2020}.

The wind- and wave-induced speed losses depend on the Beaufort number $BN$ (computed from the 10 m
wind speed; see Section~\ref{sec:data}), the Froude number, and a set of direction- and
form-dependent coefficients tabulated by \citet{yang2020}. The wind speed loss is a
fifth-order function of the wind angle relative to heading, and the wave added-resistance loss scales
with significant wave height $H_w$ and the beam-to-length ratio. The current contributes its
along-track projection, $V_{c,\parallel}=V_c\cos\theta_c$. Combining these terms gives the speed over
ground:
\begin{equation}
\label{eq:sog}
V_g \;=\; V_s - \Delta V_{\text{wind}} - \Delta V_{\text{wave}} + V_{c,\parallel},
\qquad V_g \ge 0.
\end{equation}

Two properties are used later. First, for fixed weather $V_g$ increases monotonically with $V_s$, so
the relation is invertible: a target SOG implies a unique required SWS, $V_s=g^{-1}(V_g;w)$, obtained
here by binary search (Section~\ref{sec:methods}). Second, that required SWS rises as conditions
worsen --- holding a target SOG through adverse weather demands more SWS. Both properties underlie
the convexity mechanism of Section~\ref{sec:mechanism}.

\subsection{Fuel consumption}
Fuel consumption rate is a cubic function of still-water speed \citep{yang2020}:
\begin{equation}
\label{eq:fcr}
\fcr(V_s) \;=\; 0.000706\, V_s^{3} \quad [\text{mt/h},\; V_s \text{ in kn}].
\end{equation}
This cubic form is the central structural property of the problem: \textbf{FCR is strictly convex in
$V_s$.} Crucially, fuel depends on SWS --- the speed the engine must produce --- not on SOG. When a
target SOG is held through varying weather, the required SWS varies, and the convex FCR penalises
that variation (Section~\ref{sec:mechanism}). The fuel for leg $i$ is the rate times the leg's
transit time,
\begin{equation}
\label{eq:legfuel}
F_i \;=\; \fcr(V_{s,i})\, \frac{d_i}{V_{g,i}},
\end{equation}
and total voyage fuel is the sum over legs, $F_{\text{total}}=\sum_{i=1}^{N} F_i$.

\subsection{Decision variable and objective}
This study optimises \textbf{SOG}, not SWS --- a choice termed SOG-targeting. A speed plan specifies
a target SOG for each leg (or, for the baseline, each block); the SWS required to realise it follows
from the inverse SOG relation under the prevailing weather. The optimisation problem is
\begin{equation}
\label{eq:obj}
\min_{\{V_{g,i}\}} \; \sum_{i=1}^{N} \fcr\!\big(g^{-1}(V_{g,i}; w_{i})\big)\, \frac{d_i}{V_{g,i}}
\quad\text{s.t.}\quad \sum_{i=1}^{N} \frac{d_i}{V_{g,i}} \le T, \;\; V_{s,i}\in[11,13]\ \text{kn}.
\end{equation}
The arrival constraint is hard; the still-water speed is bounded by the engine envelope. The two
formulations compared in this study (Section~\ref{sec:methods}) differ only in the granularity at
which the target SOG may be chosen --- freely per leg, or once per time block.

%% ============================================================ 4. METHODS
\section{Methods}
\label{sec:methods}
Both speed-optimisation formulations are minimum-fuel dynamic programs solved over the same route
discretisation and the same SOG grid. They differ in one respect only: the granularity of the speed
decision. The Shafir--Raviv formulation (SR) chooses speed freely on every leg; the baseline of
\citet{luo2024} (Luo) holds a single speed across each time block. Because both call the
identical SOG and fuel functions defined in Section~\ref{sec:problem} (Eqs~\ref{eq:sog},
\ref{eq:fcr}), any difference in fuel is attributable to speed-decision granularity alone.

\subsection{Common optimisation framework}
The decision variable was SOG rather than SWS: a planned SOG schedule was chosen, and the SWS
required to realise it followed from the inverse of the SOG relation under the prevailing weather
(SOG-targeting). Both formulations minimised total voyage fuel (Eq.~\ref{eq:obj}) subject to a hard
arrival-time constraint $T$. The ETA was binding in every run reported here --- arrival slack was
zero throughout --- so both formulations slow-steamed to the deadline and the comparison is one of
fuel at equal voyage time. The speed search was discretised to a common grid of 61 SOG values
spanning the mean voyage speed $\pm 3$ kn ($L/T$). The route was represented as a two-dimensional
time--distance graph with a fixed distance step and a time step $\Delta t$ \todo{state $\Delta t$ per
route}. This shared grid is the fairness control: it ensures the comparison isolates speed-decision
granularity rather than resolution.

\subsection{Free-speed formulation: SR (Shafir--Raviv)}
SR was solved on an atomic-edge graph. Each node represents a discretised (distance, time) state;
each outgoing edge represents a single speed choice over one leg, with cost equal to the leg fuel,
$c=\fcr(V_s)\,d_i/\mathrm{SOG}(V_s,w_{i,t})$, where $w_{i,t}$ is the weather encountered. The
minimum-fuel path from origin to destination within the ETA was found by a forward Bellman recursion
over the time-ordered nodes, and the optimal schedule recovered by backtracking. Because speed is
chosen independently on every leg, SR adapts to weather variation at each waypoint crossing.

\subsection{Per-block-locked baseline: Luo (2024)}
The baseline followed the block dynamic-programming formulation of \citet{luo2024}. Its graph
is a lattice of (column, distance) nodes, where columns mark times $0,\Delta t,2\Delta t,\dots$ Each
arc spans one block (one column) and is traversed at a single constant SOG equal to the block
distance divided by $\Delta t$. The arc cost was obtained by walking the weather sub-segments within
the block and summing fuel at that fixed block SOG; the minimum-fuel path to the ETA column was found
by shortest path. The defining restriction is that speed is constant within a block: the formulation
cannot vary speed in response to weather that changes inside a block. This baseline was implemented
independently from its published description --- with its own lattice and arc evaluation, not as SR
subject to an added equality constraint --- so the comparison reflects Luo's actual method rather
than a weakened variant.

\subsection{Evaluation protocol}
\label{sec:protocol}
\textbf{Consecutive-voyage chains.} Each route was evaluated as a chain in which voyage $N{+}1$
departs at the sample hour at which voyage $N$ arrives. Fixing the step to the ETA guaranteed that SR
and Luo encountered identical departure weather at every voyage. Seven voyages were run on Route~1
and twelve on Route~2 --- 19 in total --- spanning $\sim$80 days of the collection window.

\noindent\textbf{Perfect foresight (oracle).} Each leg was evaluated against the actual
weather recorded at the voyage's departure sample hour, giving perfect foresight. This bounds the
achievable advantage (Section~\ref{sec:res-oracle}).

\noindent\textbf{Rolling horizon (RH).} Both formulations were re-solved at 6 h decision steps. At
each step the first 6 h block was planned against actual (nowcast) weather and the remainder against
the most recent available predicted cycle; only the first block was committed before advancing and
re-planning. Realised fuel was the sum of the committed blocks. The 6 h cadence equals the GFS NWP
cycle (Section~\ref{sec:data}).

\noindent\textbf{Naive baseline.} As the operational reference, a set-and-forget policy sailed a
single fixed mean SOG ($L/T$) through the actual time-varying weather, with no re-planning. The
headline RH result is reported relative to this baseline (Section~\ref{sec:res-rh}).

%% ============================================================ 5. MECHANISM
\section{The convexity mechanism}
\label{sec:mechanism}
The fuel-consumption rate is strictly convex in speed: $\fcr(V_s)=0.000706\,V_s^3$ (Eq.~\ref{eq:fcr})
is cubic and therefore convex on the operating range. This single property predicts, before any
experiment, that per-block speed-locking must waste fuel relative to per-leg freedom whenever weather
varies within a block.

Consider one time block spanning several legs whose weather differs. A free-speed policy may set a
different speed on each leg; a block-locked policy must hold one speed across them all. To maintain a
constant SOG across legs of differing weather, the block-locked policy is forced to vary the
still-water speed (SWS) from leg to leg --- more SWS where conditions are adverse, less where they are
favourable. Fuel depends on SWS through the convex FCR, so by Jensen's inequality the mean fuel of
this forced, varying SWS exceeds the fuel that the mean SWS would consume:
\begin{equation}
\label{eq:jensen}
\mathbb{E}\!\left[\fcr(V_s)\right] \;\ge\; \fcr\!\left(\mathbb{E}[V_s]\right).
\end{equation}
The harsh-leg penalty always outweighs the favourable-leg saving. A free-speed policy is not bound to
a constant SOG and can allocate speed across legs so as to keep operation in a more efficient regime;
it therefore consumes no more fuel than the block-locked policy, and strictly less whenever
within-block weather is non-uniform.

Two empirical predictions follow, both tested in Section~\ref{sec:results}.
\begin{enumerate}
\item \textbf{A systematic, signed gap.} SR should consume less fuel than Luo on essentially every
voyage in which weather varies within blocks --- as a rule, because the inequality is directional.
\item \textbf{A gap that scales with weather variability.} The penalty grows with the spread of
within-block conditions; routes or departures whose weather varies more should exhibit a larger gap,
and a shorter voyage that concentrates the same variability into fewer, denser blocks should show a
proportionally larger effect.
\end{enumerate}
The same convexity has a second consequence examined later as a limitation
(Section~\ref{sec:disc-bound}): on uniform weather a constant speed is itself fuel-optimal, so the
value of any speed variation --- including a rolling-horizon policy's forecast-driven variation --- is
bounded by how much the weather actually varies. The argument is exact for the convex FCR and the
equal-time comparison used throughout; it predicts the direction and scaling of the effect, while its
magnitude on real routes is the empirical question answered next.

%% ============================================================ 6. DATA
\section{Data and experimental design}
\label{sec:data}

\subsection{Weather data}
Environmental data were obtained from the Open-Meteo API \citep{openmeteo}, which serves
operational NWP products: 10 m wind from GFS, significant wave height from MFWAM, and ocean-current
velocity from SMOC. Beaufort number was computed from the 10 m wind speed rather than taken from the
API (Section~\ref{sec:problem}). For each route, both the realised (``actual'') weather and the
forecasts (``predicted'') issued at each cycle were collected at every waypoint, at a 6 h sampling
cadence, over $\sim$80 days, and stored in HDF5 with separate tables for actual weather, predicted
weather, and metadata. The two routes differ markedly in regime (Table~\ref{tab:routes}): Route~1 is
mild and uniform; Route~2 is harsh, with mean wind and wave conditions several times larger and far
more variable \todo{insert per-route wind/wave mean $\pm$ s.d.\ from data}.

\begin{table}[ht]
\centering
\caption{Route and weather summary.}
\label{tab:routes}
\begin{tabular}{lll}
\toprule
 & Route 1 (Malacca) & Route 2 (Atlantic) \\
\midrule
Length & 3{,}393 nm & 1{,}955 nm \\
ETA & 280 h & 168 h \\
Voyages in chain & 7 & 12 \\
Wind (mean $\pm$ s.d.) & \todo{from data} & \todo{from data} \\
Wave $H_w$ (mean $\pm$ s.d.) & \todo{from data} & \todo{from data} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Forecast accuracy (supporting measurement)}
Forecast error was measured by comparing predicted weather against the actual weather subsequently
observed at the same waypoint and time, as a function of lead time. Wind-speed RMSE grew
systematically: it approximately doubled over the first 133 h on Route~1 (4.13 $\rightarrow$
8.40 km/h) and grew more steeply on Route~2 ($+286\%$ over 144 h) \todo{FIG: forecast error vs lead
time}. This degradation is why a rolling-horizon plan, which commits speeds against forecasts later
revised, consumes more fuel than the perfect-foresight oracle (Section~\ref{sec:res-rh}); it is
reported here as ground truth, independent of any optimisation.

\subsection{NWP model cycle and the re-plan cadence}
The 6 h rolling-horizon cadence was set to the underlying model refresh, not tuned. The GFS wind
product refreshes every 6 h; MFWAM (waves) and SMOC (currents) every 12 h and 24 h. Inspection of the
predicted-weather record confirmed this: 86\% of hourly wind queries returned data identical to the
previous hour. Re-planning more frequently than 6 h therefore acts on no new information.

\subsection{Experimental design}
Each route was evaluated as a consecutive-voyage chain spaced one ETA apart
(Section~\ref{sec:protocol}), yielding 19 voyages. At each departure both formulations were run under
(i) the perfect-foresight oracle and (ii) the 6 h rolling horizon, with the RH result compared
against the Naive baseline. Identical departure weather was presented to both formulations, so all
reported differences are attributable to speed-decision granularity and information regime, not
sampling.

%% ============================================================ 7. RESULTS
\section{Results}
\label{sec:results}
Throughout, SR denotes the Shafir--Raviv per-leg free-speed dynamic program and Luo the per-block
speed-locked formulation of \citet{luo2024}.

\subsection{Fuel under perfect foresight}
\label{sec:res-oracle}
Nineteen voyages were evaluated (seven on Route~1, twelve on Route~2), spanning $\sim$80 days. In
every voyage both formulations consumed the full time budget (zero arrival slack). Per-voyage fuel is
reported per voyage in Tables~\ref{tab:modec-r1} and~\ref{tab:modec-r2}; aggregates are in
Table~\ref{tab:modec}.

\begin{table}[ht]
\centering
\caption{Perfect-foresight aggregates. Negative gap = SR burns less fuel.}
\label{tab:modec}
\begin{tabular}{lrrrrr}
\toprule
Route & $n$ & SR mean $\pm$ s.d.\ (mt) & Luo mean $\pm$ s.d.\ (mt) & Mean gap (mt) & Mean gap (\%) \\
\midrule
1 (Malacca)  & 7  & $344.87 \pm 7.77$  & $351.26 \pm 8.72$  & $-6.39$ & $-1.8$ \\
2 (Atlantic) & 12 & $201.90 \pm 10.32$ & $207.36 \pm 11.03$ & $-5.46$ & $-2.6$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Perfect-foresight per-voyage fuel, Route 1 (Malacca, ETA 280 h). $\mathrm{sh}_0$ = departure sample
hour. Negative gap = SR burns less fuel.}
\label{tab:modec-r1}
\begin{tabular}{rrrrrr}
\toprule
Voyage & $\mathrm{sh}_0$ & SR (mt) & Luo (mt) & Gap (mt) & Gap (\%) \\
\midrule
0 & 6    & 354.82 & 361.56 & $-6.74$ & $-1.86$ \\
1 & 286  & 355.23 & 364.68 & $-9.45$ & $-2.59$ \\
2 & 566  & 337.70 & 342.41 & $-4.71$ & $-1.38$ \\
3 & 846  & 348.19 & 353.15 & $-4.96$ & $-1.40$ \\
4 & 1126 & 337.60 & 340.87 & $-3.27$ & $-0.96$ \\
5 & 1406 & 334.83 & 343.86 & $-9.03$ & $-2.63$ \\
6 & 1686 & 345.73 & 352.31 & $-6.58$ & $-1.87$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Perfect-foresight per-voyage fuel, Route 2 (Atlantic, ETA 168 h). $\mathrm{sh}_0$ = departure sample
hour. Negative gap = SR burns less fuel.}
\label{tab:modec-r2}
\begin{tabular}{rrrrrr}
\toprule
Voyage & $\mathrm{sh}_0$ & SR (mt) & Luo (mt) & Gap (mt) & Gap (\%) \\
\midrule
0  & 0    & 203.20 & 210.25 & $-7.05$ & $-3.35$ \\
1  & 168  & 204.17 & 209.51 & $-5.34$ & $-2.55$ \\
2  & 336  & 195.24 & 201.95 & $-6.71$ & $-3.32$ \\
3  & 504  & 206.12 & 212.22 & $-6.10$ & $-2.87$ \\
4  & 672  & 215.93 & 223.50 & $-7.57$ & $-3.38$ \\
5  & 840  & 190.75 & 196.89 & $-6.14$ & $-3.12$ \\
6  & 1008 & 227.91 & 233.85 & $-5.94$ & $-2.54$ \\
7  & 1176 & 194.40 & 198.25 & $-3.86$ & $-1.94$ \\
8  & 1344 & 194.64 & 197.05 & $-2.41$ & $-1.22$ \\
9  & 1512 & 192.75 & 196.90 & $-4.15$ & $-2.11$ \\
10 & 1680 & 199.10 & 203.49 & $-4.39$ & $-2.16$ \\
11 & 1848 & 198.57 & 204.41 & $-5.84$ & $-2.86$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{SR consumed less fuel than Luo on every voyage.} The free-speed formulation was more
fuel-efficient on all 19 of 19 voyages, with no exception on either route. The per-voyage advantage
ranged from 3.27 to 9.45 mt on Route~1 and from 2.41 to 7.57 mt on Route~2.

\noindent\textbf{The absolute advantage was comparable across routes, but proportionally larger on
the shorter voyage.} The mean saving was similar in absolute terms (6.39 mt on Route~1, 5.46 mt on
Route~2) despite Route~2 being roughly half the duration; as a fraction of total fuel the advantage
was therefore larger on the shorter, harsher Atlantic route (2.6\%) than on Malacca (1.8\%) --- the
scaling predicted by the mechanism (Section~\ref{sec:mechanism}).

\noindent\textbf{Departure weather dominated voyage-to-voyage fuel variation.} On Route~2 the worst
departure (227.91 mt) consumed 19.5\% more fuel than the best (190.75 mt) for the same formulation
--- a spread larger than the entire SR--Luo gap. The Atlantic fuel spread per voyage-hour was about
three times that of Malacca.

\subsection{Fuel under rolling-horizon planning (real forecasts)}
\label{sec:res-rh}
To test whether the advantage survives realistic operation, both formulations were embedded in the
6 h rolling-horizon scheme (Section~\ref{sec:protocol}) and compared against the Naive set-and-forget
baseline, over the same 19 voyages. Realised fuel relative to Naive is reported per voyage in
Tables~\ref{tab:rh-r1} and~\ref{tab:rh-r2} and summarised in Table~\ref{tab:rh}.

\begin{table}[ht]
\centering
\caption{Rolling-horizon realised fuel relative to the Naive set-and-forget baseline.}
\label{tab:rh}
\begin{tabular}{lrrrr}
\toprule
Route & $n$ & RH-SR vs Naive (mean \%) & RH-Luo vs Naive (mean \%) & RH-SR saves on \\
\midrule
2 (Atlantic) & 12 & $-1.9$ & $-0.2$ & 11/12 \\
1 (Malacca)  & 7  & $-1.2$ & $-0.1$ & 7/7 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Rolling-horizon per-voyage realised fuel, Route 1 (Malacca, ETA 280 h). $\mathrm{sh}_0$ =
departure sample hour. Negative \% = saving vs the Naive set-and-forget baseline.}
\label{tab:rh-r1}
\begin{tabular}{rrrrrr}
\toprule
$\mathrm{sh}_0$ & Naive (mt) & RH-SR (mt) & RH-Luo (mt) & RH-SR vs Naive (\%) & RH-Luo vs Naive (\%) \\
\midrule
6    & 362.74 & 358.86 & 362.57 & $-1.07$ & $-0.05$ \\
286  & 367.03 & 358.73 & 367.72 & $-2.26$ & $+0.19$ \\
566  & 345.42 & 342.82 & 344.51 & $-0.75$ & $-0.26$ \\
846  & 354.74 & 350.33 & 354.36 & $-1.24$ & $-0.11$ \\
1126 & 342.68 & 341.55 & 341.69 & $-0.33$ & $-0.29$ \\
1406 & 346.19 & 344.11 & 346.57 & $-0.60$ & $+0.11$ \\
1686 & 356.03 & 349.32 & 355.10 & $-1.88$ & $-0.26$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Rolling-horizon per-voyage realised fuel, Route 2 (Atlantic, ETA 168 h). $\mathrm{sh}_0$ =
departure sample hour. Negative \% = saving vs the Naive set-and-forget baseline.}
\label{tab:rh-r2}
\begin{tabular}{rrrrrr}
\toprule
$\mathrm{sh}_0$ & Naive (mt) & RH-SR (mt) & RH-Luo (mt) & RH-SR vs Naive (\%) & RH-Luo vs Naive (\%) \\
\midrule
0    & 212.61 & 205.00 & 212.12 & $-3.58$ & $-0.23$ \\
168  & 212.78 & 208.34 & 211.41 & $-2.08$ & $-0.64$ \\
336  & 203.65 & 200.99 & 203.42 & $-1.31$ & $-0.11$ \\
504  & 214.54 & 212.61 & 216.01 & $-0.90$ & $+0.68$ \\
672  & 225.98 & 222.38 & 224.90 & $-1.59$ & $-0.48$ \\
840  & 200.50 & 192.98 & 199.04 & $-3.75$ & $-0.73$ \\
1008 & 237.09 & 230.53 & 235.33 & $-2.76$ & $-0.74$ \\
1176 & 200.43 & 197.33 & 200.71 & $-1.54$ & $+0.14$ \\
1344 & 199.35 & 200.82 & 200.49 & $+0.74$ & $+0.57$ \\
1512 & 198.99 & 195.58 & 198.28 & $-1.71$ & $-0.36$ \\
1680 & 206.86 & 201.65 & 205.70 & $-2.52$ & $-0.56$ \\
1848 & 206.10 & 201.91 & 206.27 & $-2.03$ & $+0.08$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Free-speed re-planning saved fuel; block-locked re-planning did not.} Rolling-horizon
SR reduced realised fuel relative to set-and-forget on 18 of 19 voyages (mean $-1.9\%$ on Route~2,
$-1.2\%$ on Route~1; best $-3.75\%$). Rolling-horizon Luo was statistically indistinguishable from
set-and-forget (mean $-0.2\%$ and $-0.1\%$): the per-block lock left no room to exploit the refreshed
forecasts. The SR--Luo contrast established under perfect foresight therefore persisted under
realistic, imperfect-forecast operation, on both routes and across the full departure window.

\noindent\textbf{Realised fuel fell within the oracle--Naive envelope.} On every voyage the realised
RH fuel satisfied $\text{oracle}\le\text{RH}\le\text{Naive}$. Rolling-horizon SR sat between 1.8 and
8.9 mt above its oracle, the gap representing the cost of committing speeds against forecasts that
were subsequently revised.

\noindent\textbf{The two formulations re-planned differently.} A single-voyage diagnostic recorded how
often a refreshed forecast changed the committed first-block speed: rolling-horizon SR revised on 8 of
27 re-plans (30\%, mean change 0.19 kn), whereas rolling-horizon Luo revised on 17 of 27 (63\%, mean
change 0.43 kn) --- the block-locked formulation adjusted more frequently yet gained less.

\subsection{Supporting observations}
Two measurements underpin the rolling-horizon design (Section~\ref{sec:data}). First, forecast
accuracy degraded systematically with lead time (wind RMSE roughly doubled over 133 h on Route~1;
$+286\%$ over 144 h on Route~2), establishing why realised RH fuel exceeded the oracle. Second, the
6 h re-plan cadence was fixed to the GFS model cycle, of which 86\% of hourly queries returned
identical data, so sub-6 h re-planning carried no new information.

%% ============================================================ 8. DISCUSSION
\section{Discussion}
\label{sec:discussion}

\subsection{The mechanism predicted the result}
Both predictions of the convexity mechanism (Section~\ref{sec:mechanism}) were borne out. First, the
gap was signed and systematic: under perfect foresight SR consumed less fuel than Luo on all 19 of 19
voyages --- the directional signature of Jensen's inequality, not an average tendency. Second, the gap
scaled with weather variability: proportionally larger on the harsher Atlantic route (2.6\%) than on
Malacca (1.8\%), as predicted for a penalty driven by within-block weather spread. The empirical
magnitude --- roughly 6 mt per voyage --- is thus explained, not merely reported.

\subsection{Decision granularity, not data, is the limiting factor}
That the free-speed advantage persisted under realistic rolling-horizon operation
(Section~\ref{sec:res-rh}) confirms it is not an artefact of perfect information. The per-block
baseline re-planned more often under refreshed forecasts yet gained less, because it lacked the
within-block resolution to act on the new information: the limiting factor is decision granularity,
not data.

\subsection{When re-planning helps --- and when it does not}
\label{sec:disc-bound}
The rolling-horizon benefit over set-and-forget is real but bounded, and is not claimed to be
universal. On 1 of 19 voyages the free-speed RH plan consumed slightly more fuel than the fixed-speed
baseline, and the per-block RH plan did so on several. The reason is the same convexity seen from the
other side: on near-uniform weather a constant speed is itself fuel-optimal, so any forecast-driven
speed variation can only add a Jensen penalty unless the weather-routing gain exceeds it. The value of
re-planning is therefore bounded by how much the weather actually varies over a departure's window ---
largest on variable departures, vanishing or reversing on calm ones. This is best read as a
savings-versus-departure relationship \todo{FIG: savings vs departure} rather than a single figure;
the relevant gate, stated precisely, is ``realised fuel $\le$ Naive'', satisfied by SR on 18 of 19
voyages and by Luo on a minority.

\subsection{Relation to prior work}
\todo{To be written.}

\subsection{Limitations}
(i) Two routes were studied; the route-length scaling of the gap is reported as a contrast, not a
curve. (ii) The arrival constraint was hard and binding (zero slack) throughout, so results describe
fuel at equal voyage time; relaxing the ETA is left to future work. (iii) The rolling-horizon nowcast
assumes the current 6 h block is observable; its realism depends on onboard sensing. (iv) The per-block
baseline, though implemented faithfully, is one of several possible block formulations.

%% ============================================================ 9. CONCLUSION
\section{Conclusion}
\label{sec:conclusion}
This study compared per-leg free-speed voyage optimisation (the Shafir--Raviv formulation) against the
per-block speed-locked formulation of \citet{luo2024} for fuel-minimising ship routing under
time-varying weather, on two routes spanning two weather regimes.

Three conclusions follow. First, per-block speed-locking is fuel-suboptimal by a structural mechanism:
because the fuel-consumption rate is convex in still-water speed, holding one speed across a block of
varying weather forces inefficient speed excursions, and per-leg freedom recovers the loss (Jensen's
inequality). Second,
under perfect foresight the free-speed advantage was realised on every one of 19 voyages ---
comparable in absolute fuel across routes ($\sim$6 mt) but proportionally larger on the shorter,
harsher voyage. Third, the advantage survived realistic 6 h rolling-horizon operation against real
forecasts: free-speed re-planning saved fuel relative to set-and-forget on 18 of 19 voyages, whereas
per-block re-planning did not, its value bounded by weather variability.

The practical implication is that the resolution of the speed decision --- not the availability of
re-planning or fresh forecasts --- is the binding factor: a per-block planner cannot convert better
information into fuel savings, while a per-leg planner can. The expected saving can be anticipated
from the variability of the route's weather. Future work includes extending the route set
to establish the route-length scaling as a curve, relaxing the hard arrival constraint to a soft ETA
penalty, and examining the sensitivity of the conclusions to the fuel-speed exponent.

%% ============================================================ REFERENCES
\bibliographystyle{elsarticle-harv}
\bibliography{refs}

\end{document}
